Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-04-17 WSSSPE4_report/WSSSPE4_report.tex
some misc cleanup/consistency fixes

@@

All participants present on the last day of the meeting, September 14, 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
44 responses were recorded, with one respondent indicating that they took the survey twice.
The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
Hence, the total number of unique individuals responding to the survey is likely 44.
The survey contained 9 multiple choice and open-text answer questions.
Text responses have been alphabetize to preserve anonymity.
The survey had a 100\% completion rate.
Appendix~\ref{sec:survey_details} contains the detailed survey results.

@@

In WSSSPE4, we heard about a number of interesting projects and ideas, and used those
ideas to create and form working groups, intended to start at the workshop and then
continue afterwards, to address challenges that arose from the workshop ideas.  This workshop
has reinforced the lesson from WSSSPE3 that it is relatively easy to get motivated people
to attend a meeting and productively spend their time there both doing work and planning
more work, but it is very hard to get that additional work after the meeting to take place.
The main problem seems to be one of time.  Once the attendees have agree to spend
their time at the workshop, they put their energy into doing so productively, but they haven't
really committed themselves to anything more than this, so their energy and effort trails
off, as all of their other commitments (particularly, those they are funded to do as part of
their jobs) come back to the fore. Without a process to resolve this concern, the utility of
further multi-day WSSSPE workshops is unclear.

>>>>

\usepackage{placeins}

@@

All participants present on the last day of the meeting, September 14, 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
44 responses were recorded, with one respondent indicating that they took the survey twice.
The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
Hence, the total number of unique individuals responding to the survey is likely 44.
The survey contained nine multiple choice and open-text answer questions.
Text responses have been alphabetize to preserve anonymity.
The survey had a 100\% completion rate.
Appendix~\ref{sec:survey_details} contains the detailed survey results.

@@

In WSSSPE4, we heard about a number of interesting projects and ideas, and used those
ideas to create and form working groups, intended to start at the workshop and then
continue afterwards, to address challenges that arose from the workshop ideas.  This workshop
has reinforced the lesson from WSSSPE3 that it is relatively easy to get motivated people
to attend a meeting and productively spend their time there both doing work and planning
more work, but it is very hard to get that additional work after the meeting to take place.
The main problem seems to be one of time.  Once the attendees have agree to spend
their time at the workshop, they put their energy into doing so productively, but they have not
really committed themselves to anything more than this, so their energy and effort trails
off, as all of their other commitments (particularly, those they are funded to do as part of
their jobs) come back to the fore. Without a process to resolve this concern, the utility of
further multi-day WSSSPE workshops is unclear.

### 2017-04-17 WSSSPE4_report/survey_details.tex
some misc cleanup/consistency fixes

\hspace{0.2cm}

\noindent \textbf{1. Please indicate how strongly you agree or disagree with the following statements}

@@

\noindent \textbf{2. What part of the meeting was the most valuable? select all that apply}

@@

\newpage
\noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}

@@

\noindent \textbf{4. What could be improved?}

@@

\noindent \textbf{5. What would make you more likely to invest time in future WSSSPE initiatives?}

@@

\noindent \textbf{6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}

@@

\newpage
\noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}

@@

\noindent \textbf{8. Working Group Name}

\vspace{0.2cm}

@@

\noindent \textbf{9. Do you have any other comments or suggestions or ideas?}

@@

>>>>

\begin{enumerate}
\item \textbf{Please indicate how strongly you agree or disagree with the following statements}
%\noindent \textbf{1. Please indicate how strongly you agree or disagree with the following statements}

@@

%\noindent \textbf{2. What part of the meeting was the most valuable? select all that apply}
\item \textbf{What part of the meeting was the most valuable? select all that apply}

@@

%\newpage
%\noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}
\item \textbf{What additional topics would you like to see discussed at the next meeting?}

@@

%\noindent \textbf{4. What could be improved?}
\item \textbf{What could be improved?}

@@

%\noindent \textbf{5. What would make you more likely to invest time in future WSSSPE initiatives?}
\item \textbf{What would make you more likely to invest time in future WSSSPE initiatives?}

@@

%\noindent \textbf{6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}
\item \textbf{Would you be interested in joining a professional society focused around the topics of WSSSPE?}

@@

%\newpage
%\noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
\item \textbf{Will you be part of a working group with colleagues going forward on a topic addressed at WSSSPE, and if so, which one?}

@@

\FloatBarrier
%\noindent \textbf{8. Working Group Name}
\item \textbf{Working group name}

%\vspace{0.2cm}

@@

%\noindent \textbf{9. Do you have any other comments or suggestions or ideas?}
\item \textbf{Do you have any other comments or suggestions or ideas?}

@@

\end{enumerate}

### 2017-04-17 WSSSPE4_report/wg_access.tex
some misc cleanup/consistency fixes

Note that Debian provides a testing infrastructure (\href{https://ci.debian.net/doc/}{http://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.

@@

%What are the first SMART steps proposed?
\begin{enumerate}
\item Learn how to donate machines to Debian
\item Find funding, perhaps via NEIC? (\url{https://wiki.neic.no/wiki/NeIC_Community_Wiki}) 
\item Build community
\end{enumerate}

>>>>

Note that Debian provides a testing infrastructure (\url{https://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.

@@

%What are the first SMART steps proposed?
\begin{enumerate}
\item Learn how to donate machines to Debian
\item Find funding -- NeIC (\url{https://neic.nordforsk.org}) through the CodeRefinery project (\url{http://coderefinery.org}) is building a CI infrastructure (in progress) that will be provided to selected projects (through an open call)
\item Build community
\end{enumerate}

### 2017-04-17 WSSSPE4_report/wg_alliance.tex
some misc cleanup/consistency fixes

The first SMART steps of our working group are outlined below:
\begin{itemize}
\item Create a draft text to explain context and incentives for joining the Software Sustainability Alliance
\item Come up with a list of potential invitees
\item Update www.softwaresustainability.org to provide information
\item Draft a letter to go out to potential alliance members
\item Create a timeline and framework for initial consultation
\end{itemize}

@@

If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (N.ChueHong@software.ac.uk) and Jean Salac (jeansalac@virginia.edu)

>>>>

The first SMART steps of our working group are outlined below:
\begin{itemize}
\item Create a draft text to explain context and incentives for joining the Software Sustainability Alliance
\item Come up with a list of potential invitees
\item Update \url{www.softwaresustainability.org} to provide information
\item Draft a letter to go out to potential alliance members
\item Create a timeline and framework for initial consultation
\end{itemize}

@@

If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)

### 2017-04-17 WSSSPE4_report/wg_best_practices_developing.tex
some misc cleanup/consistency fixes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{White paper on developing sustainable softwares}
\label{sec:best-practices-developing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Introduction to group here, including the overall objective of work in this area.

@@

members of the working group

@@

Specific things the working group wants to accomplish in the context of the larger objective.

@@

What tasks will the working group undertake

@@

The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more information and requests, join \#wg-best-practices channel at WSSSPE's Slack team.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{White paper on developing sustainable software}
\label{sec:best-practices-developing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\note{Introduction to group here, including the overall objective of work in this area.}

@@

\note{members of the working group}

@@

\note{Specific things the working group wants to accomplish in the context of the larger objective.}

@@

\note{What tasks will the working group undertake}

@@

The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.

### 2017-04-17 WSSSPE4_report/wg_best_practices_undergrads.tex
some misc cleanup/consistency fixes

The question is, then, how to prevent researchers from writing dark
code? The participants hypothesized that the best strategy is to catch
researchers while they are still in training and teach them software
best-practices. Therefore, this working group was formed with the goal
of developing courses on software best practices aimed at
undergraduate students studying domain science. The program might be
similar to a software carpentry or data carpentry workshop but with a
focus for domain scientists.

@@

\noindent Aleksandra Nenadic and Raniere Silva are involved
in software and data carpentry.

@@

\begin{itemize}
\item Join our google group/mailing list
  \cite{WSSSPEUndergradGoogleGroup}.
\item Join the WSSSPE Slack \cite{WSSSPESlack} and sign into the
  \texttt{\#wg-undergraduatecourse} channel and say hi.
\item Ask to join the WSSSPE4-undergraduate-course organization
  \cite{WSSSPEUndergradGithub}.
\end{itemize}

>>>>

The question is, then, how to prevent researchers from writing dark
code? The participants hypothesized that the best strategy is to catch
researchers while they are still in training and teach them software
best-practices. Therefore, this working group was formed with the goal
of developing courses on software best practices aimed at
undergraduate students studying domain science. The program might be
similar to a Software Carpentry or Data Carpentry workshop but with a
focus for domain scientists.

@@

\noindent Aleksandra Nenadic and Raniere Silva are involved
in Software and Data Carpentry.

@@

\begin{itemize}
\item Join our Google Group/mailing list
  \cite{WSSSPEUndergradGoogleGroup}.
\item Join the WSSSPE Slack \cite{WSSSPESlack} and sign into the
  \texttt{\#wg-undergraduatecourse} channel and say hi.
\item Ask to join the WSSSPE4-undergraduate-course organization
  \cite{WSSSPEUndergradGithub}.
\end{itemize}

### 2017-04-17 WSSSPE4_report/wg_codemeta.tex
some misc cleanup/consistency fixes

This group, which included people who had previously been working on the CodeMeta project\footnote{\url{http://codemeta.github.io/}}, formed to address the proposed project,
 ``Define a standard for metadata to be used in current repos including authors, dependencies, ... repo url.''

>>>>

This group, which included people who had previously been working on the CodeMeta project\footnote{\url{http://codemeta.github.io/}}, formed to address the proposed project,
 ``Define a standard for metadata to be used in current repos including authors, dependencies, \dots{}, repo url.''

### 2017-04-17 WSSSPE4_report/wg_letters.tex
some misc cleanup/consistency fixes

We have identified a few next steps, even if they may not satisfy the
exact criteria of the SMART system. Specifically:
\begin{itemize}
  \item Find and review the documents listed above, send out summary
    (Wolfgang Bangerth)
  \item Gather feedback from the other members of the group
  \item Write editorial for SIAM (Wolfgang)
  \item Write 2 emails to each other group members listed above about
    also writing editorials in their communities.
\end{itemize}
In the longer term we hope to gather others in our home disciplines to
let professional organizations weigh in.

@@

The section on resources above lists a number of letters and reports
that are worth reading. Ultimately, building a community large enough
to affect change is important; contact Wolfgang Bangerth at
\url{bangerth@colostate.edu} if you are interested in helping.

>>>>

We have identified a few next steps, even if they may not satisfy the
exact criteria of the SMART system. Specifically:
\begin{itemize}
  \item Find and review the documents listed above, send out summary
    (Wolfgang Bangerth)
  \item Gather feedback from the other members of the group
  \item Write editorial for SIAM (Wolfgang)
  \item Write two emails to each other group members listed above about
    also writing editorials in their communities.
\end{itemize}
In the longer term we hope to gather others in our home disciplines to
let professional organizations weigh in.

@@

The section on resources above lists a number of letters and reports
that are worth reading. Ultimately, building a community large enough
to affect change is important; contact Wolfgang Bangerth at
<bangerth@colostate.edu> if you are interested in helping.

### 2017-04-17 WSSSPE4_report/wg_metrics.tex
some misc cleanup/consistency fixes

The gap being addressed is the lack of a standard for collecting and presenting metrics. This gap delays workflow and creates a multitude of tedious tasks, including searching for the best-fit software, unknowingly duplicating software, and other related busy-work. The need for metric standardization stems from the abundance of scientific, both `dark' and open source, software and the difficulties of ensuring the software is sustainable.

>>>>

The gap being addressed is the lack of a standard for collecting and presenting metrics. This gap delays workflow and creates a multitude of tedious tasks, including searching for the best-fit software, unknowingly duplicating software, and other related busy-work. The need for metric standardization stems from the abundance of scientific, both ``dark'' and open source, software and the difficulties of ensuring the software is sustainable.

### 2017-04-17 WSSSPE4_report/wg_open_research_index.tex
fixing typo
some misc cleanup/consistency fixes

The aim of this group is to investigate the building of an index of research products in an open sustainable manner.  Our goal is not to eliminate commercial products, but to build on what's there and provide data and services that are missing.

@@

\begin{itemize}
\item ORCID, CrossRef, DataCite, ImpactStory/depsy, altmetrics.org, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, softwareheritage.org, DBLP, eSTEP (CWO)
\item We could work with some initial publishers who don't have competing services (domain societies), PubMed
\item We need to discuss this idea with potential funders, and ask what data/services would they like to have?  And are they willing to invest in this?
\end{itemize}

@@

Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.  At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \#wg-open-research-idx.  (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

@@

\begin{enumerate}
\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people don't like, we could check what features they has and what the issues are.

@@

\item What model should be used?
- Crawling and obtaining data
- Consortium model
- Need analysis and decision

\item Need agreements with publishers, repositories, services
- Some of these will have costs, though some may be free
- Need to determine costs
- Perhaps publishers would pay us to be listed eventually, as we gain power
- To some extent, we would be competing with some of the publishers? products

\item If crawling model:
- How to crawl all of these products?
- What data (metadata) should be the output of the crawls?
    - What happens when the products being crawled and/or their metadata change?
- How to store all of this data (metadata)?

@@

\item Discuss this idea with potential partners (who, when).
Then hgt their feedback \& buy-in.
While doing so, seek suggestions for the right project manager to coordinate this

@@

If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \#wg-open-research-idx.  (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

>>>>

The aim of this group is to investigate the building of an index of research products in an open sustainable manner.  Our goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.

@@

\begin{itemize}
\item ORCID, CrossRef, DataCite, ImpactStory/Depsy, \url{altmetrics.org}, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, \url{softwareheritage.org}, DBLP, eSTEP (CWO)
\item We could work with some initial publishers who don't have competing services (domain societies), PubMed
\item We need to discuss this idea with potential funders, and ask what data/services would they like to have?  And are they willing to invest in this?
\end{itemize}

@@

Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.  At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}. (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

@@

\begin{enumerate}
\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, we could check what features they has and what the issues are.

@@

\item What model should be used?
\begin{itemize}
    \item Crawling and obtaining data
    \item Consortium model
    \item Need analysis and decision
\end{itemize}

\item Need agreements with publishers, repositories, services
\begin{itemize}
    \item Some of these will have costs, though some may be free
    \item Need to determine costs
    \item Perhaps publishers would pay us to be listed eventually, as we gain power
    \item To some extent, we would be competing with some of the publishers? products
\end{itemize}

\item If crawling model:
\begin{itemize}
    \item How to crawl all of these products?
    \item What data (metadata) should be the output of the crawls? What happens when the products being crawled and/or their metadata change?
    \item How to store all of this data (metadata)?
\end{itemize}

@@

\item Discuss this idea with potential partners (who, when).
Then get their feedback \& buy-in.
While doing so, seek suggestions for the right project manager to coordinate this

@@

If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \texttt{\#wg-open-research-idx}. (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

### 2017-04-17 WSSSPE4_report/wg_prototyping.tex
some misc cleanup/consistency fixes

This group attempts to tackle the latter challenge through the development of
novel software prototyping infrastructure with an emphasis at a higher degree
of abstraction. The group deems the ability to rapidly construct software artifacts that can be
trusted in terms of the methods from their design up, easily discarded when
wrong and extended when right at low human and computational cost to be one
aspect of scientific software sustainability. In essence,
a motto for scientific software prototyping is {\it fail hard, fail fast}.

@@

In terms of practice and experience, a sample scientific community is required
in order to work around the capabilities of the tool. In addition, use cases
are required (e.g. micromagnetics~\cite{fischbacher2007systematic}).

>>>>

This group attempts to tackle the latter challenge through the development of
novel software prototyping infrastructure with an emphasis at a higher degree
of abstraction. The group deems the ability to rapidly construct software artifacts that can be
trusted in terms of the methods from their design up, easily discarded when
wrong and extended when right at low human and computational cost to be one
aspect of scientific software sustainability. In essence,
a motto for scientific software prototyping is \textit{fail hard, fail fast}.

@@

In terms of practice and experience, a sample scientific community is required
in order to work around the capabilities of the tool. In addition, use cases
are required (e.g., micromagnetics~\cite{fischbacher2007systematic}).
